---
title: "Untitled"
output: html_document
date: "2024-11-05"
---
# 0. Contribution Statement
Student 1:

Student 2: 
# 1. Introduction
This study investigates the distribution of palindromic sequences within the DNA structure of the cytomegalovirus (CMV), a virus known for its complex genome. Specifically, it focuses on understanding whether these palindromic sequences—sections of DNA that read the same forwards and backwards—are randomly scattered or form patterns that may indicate underlying biological functions. The CMV DNA consists of 229,354 base pairs, with researchers identifying 296 palindromic sequences at least 10 pairs long, located across various positions within the genome.

The main objective of this study is to analyze the structural distribution of these palindromic sites, assessing if their arrangement deviates from a uniform scatter across the DNA sequence. Identifying clusters in this distribution may provide clues to essential viral mechanisms, such as origins of replication. Data from this study is derived from a dataset containing the positions of these 296 palindromes.

1. Does the distribution of palindromic sites resemble a random scatter across the DNA sequence?
2. What patterns, if any, exist in the spacing between consecutive palindromic sites, and do these patterns suggest any underlying structure?
3.
4.



```{r}
data <- read.table('hcmv.txt',header = TRUE)
data
```
## Question 1
### Method

### Analysis
```{r}
num_simulations <- 10
num_palindromes <- 296

sequence_length <- 229354


# List to store simulated sets
simulated_data <- list()

# Loop to generate multiple simulated sets
for (i in 1:num_simulations) {
  simulated_data[[i]] <- sample(1:sequence_length, 296, replace = FALSE)
}
```


```{r}
calculate_distances <- function(data) {
  sorted_data <- sort(data)
  diff(sorted_data)
}

# Calculate distances for real data
sorted_data <- sort(data$location)
real_distances <- diff(sorted_data)

# Calculate distances for each simulated set
simulated_distances <- lapply(simulated_data, calculate_distances)

# Plot the real and simulated data
num_simulations <- 10
num_palindromes <- 296

sequence_length <- 229354


# List to store simulated sets
simulated_data <- list()

# Loop to generate multiple simulated sets
for (i in 1:num_simulations) {
  simulated_data[[i]] <- sample(1:sequence_length, 296, replace = FALSE)
}
```


```{r}
calculate_distances <- function(data) {
  sorted_data <- sort(data)
  diff(sorted_data)
}

# Calculate distances for real data
sorted_data <- sort(data$location)
real_distances <- diff(sorted_data)

# Calculate distances for each simulated set
simulated_distances <- lapply(simulated_data, calculate_distances)
simulated_distances
```


```{r}

# Plot the real and simulated data
plot(data$location, rep(1, num_palindromes), pch = 16, col = "blue",
     main = "Real vs. Simulated Palindrome Locations",
     xlab = "Position along DNA sequence", ylab = "Simulation Index",
     ylim = c(0.5, num_simulations + 1))

# Plot simulated palindrome locations
for (i in 1:num_simulations) {
  points(simulated_data[[i]], rep(i + 1, num_palindromes), pch = 16, col = "red")
}
```

## Question 2
### a. 
```{r}
hist(real_distances, breaks = 20, col = "blue", main = "Spacing between Consecutive Palindromes (Real)",
     xlab = "Spacing", xlim = c(min(real_distances), max(unlist(simulated_distances))),
     ylab = "Frequency", freq = TRUE)

# Add title for clarity
title(main = "Spacing between Consecutive Palindromes (Real)")

# Now plot the spacing for one simulated set
hist(simulated_distances[[1]], breaks = 20, col = rgb(1, 0, 0, 0.5), main = "Spacing between Consecutive Palindromes (Simulated)",
     xlab = "Spacing", xlim = c(min(real_distances), max(unlist(simulated_distances))),
     ylab = "Frequency", freq = TRUE)

# Add title for clarity
title(main = "Spacing between Consecutive Palindromes (Simulated)")

```
### 2b
```{r}
palindrome_pairs_sum <- rowSums(cbind(sorted_data[-length(sorted_data)], sorted_data[-1]))
head(palindrome_pairs_sum)
# Create a histogram for the sum of palindrome pairs
hist(palindrome_pairs_sum, breaks = 20, col = "blue", main = "Sum of Palindrome Pairs", xlab = "Sum of Consecutive Pairs", ylab = "Frequency")


```
### 2c
```{r}
# Ensure palindrome locations are sorted (real data)
sorted_locations <- sort(data$location)  # Assuming 'data$location' contains your palindrome positions

# Ensure that the number of palindrome locations is a multiple of 3
num_palindromes <- length(sorted_locations)

# If the number of palindromes is not a multiple of 3, adjust by removing extra locations
num_triplets <- floor(num_palindromes / 3) * 3  # Adjust to be a multiple of 3
sorted_locations <- sorted_locations[1:num_triplets]  # Trim excess if necessary

# Create triplets by grouping consecutive locations
# We will use seq() to generate triplets
triplets <- cbind(sorted_locations[seq(1, num_triplets-2, by=3)],   # First element of each triplet
                  sorted_locations[seq(2, num_triplets-1, by=3)],   # Second element of each triplet
                  sorted_locations[seq(3, num_triplets, by=3)])     # Third element of each triplet

# Calculate the sum of each triplet
palindrome_triplets_sum <- rowSums(triplets)

# View the first few sums of triplets
head(palindrome_triplets_sum)

# Simulate random palindrome locations
set.seed(123)  # Set seed for reproducibility
num_simulations <- 1000  # Number of random simulations
random_simulations <- list()

for (i in 1:num_simulations) {
  simulated_locations <- sample(1:229354, num_palindromes, replace = FALSE)  # Random sample from DNA sequence
  sorted_simulated_locations <- sort(simulated_locations)
  
  # Ensure the number of simulated locations is a multiple of 3
  num_sim_triplets <- floor(length(sorted_simulated_locations) / 3) * 3
  sorted_simulated_locations <- sorted_simulated_locations[1:num_sim_triplets]  # Trim if necessary
  
  # Create triplets for the simulated data
  simulated_triplets <- cbind(sorted_simulated_locations[seq(1, num_sim_triplets-2, by=3)],   # First element of each triplet
          sorted_simulated_locations[seq(2, num_sim_triplets-1, by=3)],   # Second element of each triplet
          sorted_simulated_locations[seq(3, num_sim_triplets, by=3)])     # Third element of each triplet
  # Calculate the sum of each triplet in the simulated data
  simulated_triplets_sum <- rowSums(simulated_triplets)
  # Store the results
  random_simulations[[i]] <- simulated_triplets_sum
}

# Check the first few sums of simulated triplets (for one simulation)
head(random_simulations[[1]])

# Create a histogram for the observed sums of triplets (real data)
# Create a histogram for the observed sums of triplets (real data)
hist(palindrome_triplets_sum, breaks = 30, col = rgb(0, 0, 1, 0.5),
     main = "Observed Sum of Palindrome Triplets",
     xlab = "Sum of Consecutive Triplets", ylab = "Frequency")

# Create separate histograms for the simulated sums of triplets (one for each simulation)
par(mfrow = c(2, 2))  # Set up a 2x2 grid for multiple plots (adjust as needed)

for (i in 1:4) {
  hist(random_simulations[[i]], breaks = 30, col = rgb(1, 0, 0, 0.5),
       main = paste("Simulated Sum of Triplets (Simulation", i, ")"),
       xlab = "Sum of Consecutive Triplets", ylab = "Frequency")
}

# Reset the plotting layout
par(mfrow = c(1, 1))  # Reset to the default single plot layout



```


### Question 3

```{r}



```

